One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2021-09-10 paper.tex
edits up to 6.2

The predicted mean of the posterior intensity field is shown in \autoref{fig:intensity-mean-cv}A along with a map of the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid with each prediction cell area approximately 1.7 hectares.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity.  There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north.  In the north, estimated intensities are lower, which also contributes to larger CV values and makes the CV map hard to interpret.  The standard deviation map shows stronger overall posterior variability in the south where intensity is larger, most likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.

@@

The most common method to communicate uncertainty in a maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} that can be derived from the posterior predictive distribution for the model or bootstrapping in a maximum likelihood context.  Another approach is to map quantiles or the probability of exceeding certain thresholds for each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.  In Section \ref{sec-results}, we present maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of the posterior intensity field that are all intended to communicate uncertainty in the spatial distribution of \akepa{} (Figures \ref{fig:intensity-mean-cv}B, \ref{fig:intensity-mean-cv}C, \ref{fig:intensity-quantiles}A, and \ref{fig:intensity-quantiles}B, respectively).  Whilst maps that summarise spatial predictions are useful, they all mask certain properties that may be important when communicating the results of the analysis.

A map showing the posterior predicted mean across the study region (Figure \ref{fig:intensity-mean-cv}A) is often the key output of a species distribution model.  However, even this relatively innocuous summary statistic can mask important features of the random field since the posterior mean will always be smoother than realisations of the field.
\begin{figure}[!htb]
	\includegraphics[scale=0.525]{figures/intensity_realized.png}
	\caption{Three realizations of the posterior intensity field}
	\label{fig:intensity-realizations}
\end{figure}
\autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.  Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).  Hence, considering this finer-grained structure, our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the map of the posterior mean.  This is one example of how a summary statistic can mask important features of the posterior random field.

For model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data and thus what we might encounter in the field.  The mean is only a summary of the posterior intensity field and can give a misleading sense of homogeneity, particularly in the high density area in the south, when compared to realisations of the mean.  Presenting multiple realisations, perhaps as an animation, is also an effective way to communicate uncertainty. \cite{bowman_GraphicsUncertainty_2019a} provide an interpolation method that preserves the mean and covariance structure of realisations that smoothly interpolates between realisations.  This avoids abrupt changes that can occur when animating the raw realisations.

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  The observed higher CV areas in such maps will highlight regions of relatively lower intensity but not necessarily higher overall uncertainty.  This is clearly the case in \autoref{fig:intensity-mean-cv}B.  However, there is also variable sampling effort within the study region and it is not clear what impact this has on the CV map.  This means that the resulting map is hard to interpret.

The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values.  However, care should be taken as to what the mapped values imply in the context of a given analysis.  A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  What such a map does not show is whether these differences actually matter in the context of the analysis.  For example, it could be the case that the standard deviation values are so large as to make the predictions have an unacceptably high uncertainty everywhere.  Or, on the other hand, perhaps differences between high and low uncertainty regions (where `high' and `low' are defined by the default colour scale for the map) are small enough to be negligible when it comes to making conclusions or decisions, in which case such a spatially varying map may not be a useful output to communicate to decision makers.  These are two extreme examples.  However, even in less extreme situations, the actual practical relevance of the SD values should be considered in order to be able to interpret the map.

The SD map suffers from another weakness which is that the SD values are linked to the predicted mean since the model assumes a log-Gaussian relationship.  This means that the SD map usually looks similar to the mean intensity map.  This relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.  One might expect uncertainty to be lower in areas with greater survey effort, however for the \akepa{} data the opposite is the case.  This is another limitation of the SD map.

The quantile maps (Figure \ref{fig:intensity-quantiles}) are also potentially difficult to interpret. The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data with one possible intensity surface (0.025 quantile map) that signifies lower abundance and another surface (0.975 quantile map) that signifies higher abundance.  However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction locations to \textit{simultaneously} achieve their 0.025 or 0.975 quantiles.  A similar problem will occur for maps of the marginal probability of exceeding a threshold at each prediction location.

Presenting independent quantiles jointly in a single image is risky.  We believe these caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps if they are not careful.  To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.  A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance. However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).  This demonstrates that the tendency to interpret these maps as showing possible intensity surfaces that are consistent with the observed data can lead to interpretations that are inconsistent with the very model that generated the maps.

We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret.  We next present some alternatives that avoid some, but not all, of these problems.

>>>>

The predicted mean of the posterior intensity field is shown in \autoref{fig:intensity-mean-cv}A along with a map of the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid with each prediction cell area approximately 1.7 hectares.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity.  There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north.  In the north, estimated intensities are lower, which also contributes to larger CV values and makes the CV map hard to interpret.  The standard deviation map shows stronger overall posterior variability in the south where intensity is larger, most likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.

@@

The most common method to communicate uncertainty in a maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}.  These can be derived from the posterior predictive distribution for the model in a Bayesian framework and bootstrapping in a maximum likelihood context.  Another approach is to map quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.  In Section \ref{sec-results}, we present maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of the posterior intensity field (Figures \ref{fig:intensity-mean-cv}B, \ref{fig:intensity-mean-cv}C, \ref{fig:intensity-quantiles}A, and \ref{fig:intensity-quantiles}B, respectively) that are all intended to communicate uncertainty in the estimate of the \akepa{} spatial distributions.  Whilst such maps that summarise spatial predictions are useful, they all mask certain properties that may be important when communicating the results of the analysis.

A map showing the posterior predicted mean across the study region (Figure \ref{fig:intensity-mean-cv}A) is often the key output of a species distribution model.  However, the posterior mean will always be smoother than realisations of the field itself, highlighting that even this relatively innocuous summary statistic can mask important features of the random field.  \autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.  Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).  Our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the map of the posterior mean.
\begin{figure}[!htb]
	\includegraphics[scale=0.525]{figures/intensity_realized.png}
	\caption{Three realizations of the posterior intensity field}
	\label{fig:intensity-realizations}
\end{figure}

For model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data and thus what ecologists might encounter in the field.  The mean can give a sense of homogeneity, particularly in the high density area in the south, when compared to realisations of the mean.  Presenting multiple realisations, perhaps as an animation, is also an effective way to communicate uncertainty. \cite{bowman_GraphicsUncertainty_2019a} provide an interpolation method that preserves the mean and covariance structure of realisations that smoothly interpolates between realisations.  This avoids abrupt changes that can occur when animating the raw realisations.

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will be generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.  However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.  It is not clear what impact this has on the CV values and this makes the map hard to interpret.

The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.  However, care should be taken as to what the mapped values imply in the context of a given analysis.  A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  However, whether these relative differences matter is dependent on the overall context of the analysis.  Some levels of variability may be acceptable in some contexts and unacceptable in others.

The SD values also reflect a modelling \textit{assumption} but not necessarily variability due to varying survey effort.  The SD is linked to the predicted mean by the assumed log-Gaussian relationship which means the mean and variance are directly connected, with higher mean values necessitating higher SD values.  This means that the SD map will produce a pattern similar to the mean intensity map, as is the case here (Figures 2A and 2B).  This relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.  One might expect uncertainty to be lower in areas with greater survey effort, however for the \akepa{} data the opposite is the case when interpreting the SD map.

The quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret. The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data. One possible intensity surface (0.025 quantile map) that signifies a kind of lower bound on abundance and another surface (0.975 quantile map) that signifies higher abundance.  However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.  A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location.

Presenting independent quantiles jointly in a single image is risky.  These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.  To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.  A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance. However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).  This shows that the tendency to interpret these maps as showing possible intensity surfaces can lead to interpretations that are inconsistent with the very model that generated the maps.

We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret. In the following suggestion we suggest an approach that avoids some, but not all, of the pitfalls above.
